Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-17 Paper_Draft2/Draft2.tex

% Authors, for the paper (add full first names)
\Author{Amrita Acharya$^{1, 2}$, Dianne Caravela$^{1, 3}$, Eunice
Kim$^{1, 4}$, Emma Kornberg$^{1, 5}$, Elisabeth Nesmith$^{1, 6}$}

@@

\caption{This plot shows the distribution of two year recidivism outcomes. Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2600 did not.}\label{fig:recid plot}
\end{figure}

@@

\begin{longtable}[]{@{}lll@{}}
\caption{This table shows the counts of recidivism by race, illustrating
how a much greater proportion (\textgreater{} 50\%) of Black defendants
recidivated than their white counterparts.
\label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived \\
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived \\
\midrule
\endhead
African American & 1661 & 1512 \\
Caucasian & 822 & 1278 \\
\bottomrule
\end{longtable}

@@

@@

\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
the favorable outcome occurs in one group versus the other. In the case
of recidivism, this is the ratio of how many white defendants are
predicted to not recidivate compared to how many Black defendants are
predicted to not recidivate. A value of 1 means that the ratio is
exactly 1:1. Less than 1 means the privileged group (white defendants)
benefits, while a value greater than 1 means the unprivileged group
(Black defendants) benefits. According to AIF360, a ratio between .8 to
1.25 is considered fair:

@@

\textbf{Average Odds Difference} - This metric returns the average
difference in false positive rate and true positive rate for the
privileged and unprivileged groups. A value of 0 indicates equality of
odds, and a value below 0 implies benefit for the privileged group.
Equality of odds is achieved in the case of recidivism when the
proportion of people who were predicted to recidivate and did recidivate
is equal (true positive rate) for both Black and white defendants AND
the proportion of people who were predicted to recidivate and did not
recidivate (false positive rate) is equal for both Black and white
defendants.

@@

\textbf{Equal Opportunity Difference} - This metric is computed as the
difference of true positive rates between the unprivileged and the
privileged groups. The true positive rate is the ratio of true positives
to the total number of actual positives for a given group.

The ideal value is 0. A value less than 0 implies higher benefit for the
privileged group and a value greater than 0 implies higher benefit for
the unprivileged group. Fairness for this metric is between -0.1 and
0.1. This metric is best used when it is very important to catch
positive outcomes while false positives are not exceptionally
problematic. This is not the case for the COMPAS data set, as false
positives mean extra jail time for someone who will not actually
re-offend.

@@

For the next step of our experiment, we needed to determine where in the
data science pipeline we can mitigate the most bias, using
pre-processing, in-processing, and post-processing de-biasing
algorithms. These are all based on using predictive models to figure out
how we can ``fix'' the bias that is present. Pre-processing refers to
assessing the training data, and it is the most flexible method because
it has not yet trained a model that may carry assumptions about the
data. It is important to keep in mind that pre-processing prevents
assumptions in the modeling, but does not account for the bias in data
collection. With pre-processing, we use the re-weighing algorithm from
AIF360 which weights the examples in each (group, label) combination
differently to ensure fairness before classification
\citep{aif360-oct-2018}. We used a logistic regression model for this
algorithm, as it is the easiest to interpret in the given context. After
running the fairness metrics using the pre-processing algorithm, we were
able to compare our results to the baseline metrics from the previous
section.

Another approach, in-processing, explores how bias can be targeted while
building a model. It assesses how we build learning algorithms, which
often prioritize accuracy over fairness. The technique we use is the
Meta Classification algorithm, which accounts for the fairness metric as
part of the input and returns a classifier optimized by that particular
metric. In order to do this, we first needed to convert our data frame
into a data type called a . Similarly to pre-processing, we compared the
results of our in-processing methods with both the baseline and the
pre-processing to gauge which method so far has better individual or
group fairness.

Our last approach, post-processing, makes transformations on model
outputs. Similarly to the pre-processing, it is flexible because it does
not work directly with the algorithm, but instead has direct access to
all the predicted classification values. Because the algorithms are
often so complex and difficult to interpret, post-processors are trained
in a black box. To get around this, we use the PostProcessingMeta class
from AIF360 which combines the training and prediction for a random
estimator and the post-processor while simultaneously splitting the data
set. Based on these results, we looked comparatively at all the
approaches to determine which we assessed to be the most fair.

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

>>>>

% Authors, for the paper (add full first names)
\Author{Amrita Acharya$^{}$, Dianne Caravela$^{}$, Eunice Kim$^{}$, Emma Kornberg$^{}$, Elisabeth Nesmith$^{}$}

@@

\caption{This plot shows the distribution of two year recidivism outcomes. Out of all defendants, there is a higher proportion of people who did not recidivate than who did. About 2500 people recidivated, whereas approximately 2700 did not.}\label{fig:recid plot}
\end{figure}

@@

\begin{longtable}[]{@{}lll@{}}
\caption{This table shows the counts of recidivism by race, illustrating
how a much greater proportion (\textgreater{} 50\%) of Black defendants
recidivated than their white counterparts.
\label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived\tabularnewline
\midrule
\endhead
African American & 1661 & 1512\tabularnewline
Caucasian & 822 & 1278\tabularnewline
\bottomrule
\end{longtable}

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to
``to help facilitate the transition of fairness research algorithms to
use in an industrial setting and to provide a common framework for
fairness researchers to share and evaluate algorithms''
\citep{aif360-oct-2018}. It contains multiple data sets, including the
COMPAS data set that accompanied the Angwin et al.~piece.

@@

\textbf{Disparate Impact Ratio} - This metric is the ratio of how often
the favorable outcome occurs in one group versus the other. In the case
of recidivism, this is the ratio of how many white defendants are
predicted to not recidivate compared to how many Black defendants are
predicted to not recidivate. A value of 1 means that the ratio is
exactly 1:1. Less than 1 means the privileged group (white defendants)
benefits, while a value greater than 1 means the unprivileged group
(Black defendants) benefits. According to AIF360, a ratio between .8 to
1.25 is considered fair\citep{Ronaghan2019AI}.

@@

\textbf{Average Odds Difference} - This metric returns the average
difference in false positive rate and true positive rate for the
privileged and unprivileged groups. A value of 0 indicates equality of
odds, and a value below 0 implies benefit for the privileged group.
Equality of odds is achieved in the case of recidivism when the
proportion of people who were predicted to recidivate and did recidivate
is equal (true positive rate) for both Black and white defendants AND
the proportion of people who were predicted to recidivate and did not
recidivate (false positive rate) is equal for both Black and white
defendants\citep{aif360-oct-2018}.

@@

\textbf{Equal Opportunity Difference} - This metric is computed as the
difference of true positive rates between the unprivileged and the
privileged groups. The true positive rate is the ratio of true positives
to the total number of actual positives for a given
group\citep{GoogleDev}.

The ideal value is 0. A value less than 0 implies higher benefit for the
privileged group and a value greater than 0 implies higher benefit for
the unprivileged group. Fairness for this metric is between -0.1 and 0.1
\citep{aif360-oct-2018}.

This metric is best used when it is very important to catch positive
outcomes while false positives are not exceptionally problematic
\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
false positives mean extra jail time for someone who will not actually
re-offend.

@@

For the next step of our experiment, we need to determine where in the
data science pipeline we can mitigate the most bias, using
pre-processing, in-processing, and post-processing de-biasing
algorithms. These are all based on using predictive models to figure out
how we can ``fix'' the bias that is present.

Pre-processing refers to assessing the training data, and it is the most
flexible method because it has not yet trained a model that may carry
assumptions about the data. It is important to keep in mind that
pre-processing prevents assumptions in the modeling, but does not
account for the bias in data collection. With pre-processing, we use the
re-weighing algorithm from AIF360 which weights the examples in each
(group, label) combination differently to ensure fairness before
classification \citep{aif360-oct-2018}. We use a logistic regression
model for this algorithm, as it is the easiest to interpret in the given
context. After running the fairness metrics using the pre-processing
algorithm, we are able to compare our results to the baseline metrics
from the previous section.

Another approach, in-processing, explores how bias can be targeted while
building a model. It assesses how we build learning algorithms, which
often prioritize accuracy over fairness. The technique we use is the
Meta Classification algorithm, which accounts for the fairness metric as
part of the input and returns a classifier optimized by that particular
metric. Similarly to pre-processing, we compare the results of our
in-processing methods with both the baseline and the pre-processing to
gauge which method so far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model
outputs. Similarly to the pre-processing, it is flexible because it does
not work directly with the algorithm, but instead has direct access to
all the predicted classification values. Because the algorithms are
often so complex and difficult to interpret, post-processors are trained
in a black box. To get around this, we use the PostProcessingMeta class
from AIF360 which combines the training and prediction for a random
estimator and the post-processor while simultaneously splitting the data
set \citep{aif360-oct-2018}. Based on these results, we look
comparatively at all the approaches to determine which we assess to be
the most fair.

@@

\begin{longtable}[]{@{}llll@{}}
\caption{Results of Baseline Fairness Metric Analysis
\label{tab:baseline metrics table}}\tabularnewline
\toprule
Fairness Metric & Ideal Value & Baseline Value & Benefited
Group\tabularnewline
\midrule
\endfirsthead
\toprule
Fairness Metric & Ideal Value & Baseline Value & Benefited
Group\tabularnewline
\midrule
\endhead
Statistical Parity Difference & 0 & -0.14 & White
Defendants\tabularnewline
Disparate Impact Ratio & 1 & 0.47 & White Defendants\tabularnewline
Average Odds Difference & 0 & -0.44 & White Defendants\tabularnewline
Equal Opportunity Difference & 0 & -0.41 & White
Defendants\tabularnewline
\bottomrule
\end{longtable}
